Technique for \earlywarning\ detection of compact binary coalescence and its implications for multi-messenger astronomy

### 2011-06-10 method.tex
Reorganize section 2.3 to make happy grad student happier.
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@1012 ccdd05d5-56c1-461a-bf98-85f6876523b4

The first step of our proposed method is to divide the templates into time
slices in a \TD{} analogue to the \FD{} decomposition described
in~\citep{Marion2004, Buskulic2010}.  To wit,
%
\begin{equation}
\label{eq:time-slices}
h_{i}[k] = \sum_{s=0}^{S-1}
	\begin{cases}
		h_i^s[k] & \textrm{if } t^s \leqslant k / f^0 < t^{s+1} \\
		0 & \textrm{otherwise}
	\end{cases}
\end{equation}
%
for $S$ integers $\{f^0 t^s\}$ such that $0  = f^0 t^0 < f^0 t^1 < \cdots < f^0
t^S = N$.  A matched filter is constructed for each time slice.  The outputs
form an ensemble of partial \SNR{} streams.  By linearity, these partial \SNR\
streams can be suitably time-delayed and summed to reproduce the \SNR\ of the
full template.  We will show in the next section that this, combined with the
\SVD{}, is sufficient to enable a computationally efficient \TD{} search and
furthermore is an essential part of an \earlywarning\ detection scheme.

For concreteness and simplicity, we will consider an inspiral waveform in the
quadrapole approximation, for which the time-frequency relation is
%
\begin{equation} \label{eq:fgw}
%
f = \frac{1}{\mathcal{\pi M}} \left[ \frac{5}{256}\frac{\mathcal{M}}{t}
\right]^{3/8}.
%
\end{equation}
%
Here, $\mathcal{M}$ is the chirp mass of the binary in units of time (where $G
M_\odot / c^3 \approx 5 \upmu\mathrm{s}$) and $t$ is the time relative to the
coalescence of the binary~\citep{findchirppaper, kidder1992}.  Usually the
template is truncated at some prescribed time $t^0$, or equivalently frequency
$f_\mathrm{hi}$.  This is often chosen to correspond to the \ISCO. An inspiral
signal will enter the detection band at a low frequency, $f = f_\mathrm{low}$,
corresponding to a time $t_\mathrm{low}$.  The template is assumed to be zero
outside the interval $[t^0, t_\mathrm{low})$ and is said to have  a duration of
$t_\mathrm{low} - t^0$. It is critically sampled at a rate of $2
f_\mathrm{hi}$.

The monotonic time-frequency relationship of equation~\eqref{eq:fgw} allows us
to choose time-slice boundaries that require substantially less bandwidth at
early times in the inspiral.  Our goal is to reduce the filtering cost of a
large fraction of the waveform by computing part of the convolution at a lower
sample rate.  Specifically we consider here time slice boundaries with the
smallest power-of-two sample rate that sub-critically samples the time-sliced
template.  The time slices consist of the $S$ intervals
$\left[t^0, t^1\right),\, \left[t^1, t^2\right),\, \dots,\, \left[t^{S-1}, t^S\right)$,
sampled at frequencies $f^0,\, f^1,\, \dots,\, f^\mathrm{S-1}$ where $f^0
\geqslant 2 f_\mathrm{hi}$ and $f^{S-1} \geqslant 2 f_\mathrm{low}$.  The time
sliced templates may be down-sampled without aliasing, so we define them as
%
\begin{equation}
\label{eq:time-sliced-templates}
h_{i}^{s}[k] \equiv
	\begin{cases}
		h_{i}\!\left[k\frac{f}{f^s}\right] & \textrm{if } t^s \leqslant k/f^s < t^{s+1} \\
		0 & \textrm{otherwise.}
	\end{cases}
\end{equation}
%
\begin{comment}
An example time-slice design satisfying these constraints for a $1.4 - 1.4 \,
M_{\odot}$ binary is shown in table~\ref{table:time_slices}.
%
\begin{table}[h!]
\begin{minipage}[c]{0.52\textwidth}
\centering
\vspace{0.8cm}
\includegraphics{time_slices.pdf}
\end{minipage}
\begin{minipage}[c]{0.3\textwidth}
\centering
\input{time_slices.tex}
\end{minipage}
\caption{\label{table:time_slices} Example of nearly critically sampled,
power-of-two time slices for a $1.4 - 1.4 \, M_{\odot}$ template extending from
$f_\mathrm{low} = 10 \, \mathrm{Hz}$ to $f_\mathrm{hi} = f_\mathrm{ISCO} = 1571\, \mathrm{Hz}$
with a time-frequency structure given by ($\ref{eq:fgw})$. $f^s$ is the sample
rate of the time slice, $(t^{s+1}, t^s]$ are the boundaries in seconds
preceding coalescence and \slicessamps\ is the number of samples in the
$s^{\mathrm{th}}$ filter.}
\end{table}
\end{comment}

Since waveforms with neighboring intrinsic source parameters $\mathbf\Theta$ have
similar time-frequency evolution, it is possible to design computationally
efficient time slices for an extended region of parameter space rather than to
design different time slices for each template.

We note that the time-slice decomposition in equation~(\ref{eq:time-slices}) is
manifestly orthogonal since the time slices are disjoint in time.  In the next
section we examine how to reduce the number of filters within each time slice
via \SVD{} of the time-sliced templates.

>>>>

The first step of our proposed method is to divide the templates into time
slices in a \TD{} analogue to the \FD{} decomposition described
in~\citep{Marion2004, Buskulic2010}.  To wit, we decompose each template
$h_{i}[k]$ into a sum of $S$ non-overlapping filters
%
\begin{equation}
\label{eq:time-slices}
h_{i}[k] = \sum_{s=0}^{S-1}
	\begin{cases}
		h_i^s[k] & \textrm{if } t^s \leqslant k / f^0 < t^{s+1} \\
		0 & \textrm{otherwise}
	\end{cases}
\end{equation}
%
for $S$ integers $\{f^0 t^s\}$ such that $0  = f^0 t^0 < f^0 t^1 < \cdots < f^0
t^S = N$.  A matched filter is constructed for each time slice.  The outputs
form an ensemble of partial \SNR{} streams.  By linearity of the filtering
process, these partial \SNR\ streams can be suitably time-delayed and summed to
 reproduce the \SNR\ of the full template.

Since waveforms with neighboring intrinsic source parameters $\mathbf\Theta$
 have similar time-frequency evolution, it is possible to design computationally
efficient time slices for an extended region of parameter space rather than to
design different time slices for each template.
For concreteness and simplicity, consider an inspiral waveform in the
quadrapole approximation, for which the time-frequency relation is
%
\begin{equation} \label{eq:fgw}
%
f = \frac{1}{\mathcal{\pi M}} \left[ \frac{5}{256}\frac{\mathcal{M}}{t}
\right]^{3/8}.
%
\end{equation}
%
Here, $\mathcal{M}$ is the chirp mass of the binary in units of time (where $G
M_\odot / c^3 \approx 5 \upmu\mathrm{s}$) and $t$ is the time relative to the
coalescence of the binary~\citep{findchirppaper, kidder1992}.
The monotonic time-frequency relationship of equation~\eqref{eq:fgw} allows us
to choose time-slice boundaries that require substantially less bandwidth at
early times in the inspiral.

In any given interval of time, the filters are band-limited across the enitre
template bank. Usually the template is truncated at some prescribed time $t^0$,
or equivalently frequency $f_\mathrm{hi}$, often chosen to correspond to the
\ISCO. An inspiral signal will enter the detection band at some low frequency,
$f = f_\mathrm{low}$, corresponding to a time $t_\mathrm{low}$.  The end of the
template is then critically sampled at a rate of $2 f_\mathrm{hi}$ but the 
beginning of the template is critically sampled at $2 f_\mathrm{lo}$. In any
time interval smaller than the duration of the template, the bandwidth of the
filters across the entire template bank may be significantly less and
a lower sampling rate may be used for all filters without aliasing.

Our goal is to reduce the filtering cost of a
large fraction of the waveform by computing part of the convolution at a lower
sample rate.  Specifically we consider here time slice boundaries with the
smallest power-of-two sample rate that sub-critically samples the time-sliced
template.  The time slices consist of the $S$ intervals
$\left[t^0, t^1\right),\, \left[t^1, t^2\right),\, \dots,\, \left[t^{S-1}, t^S\right)$,
sampled at frequencies $f^0,\, f^1,\, \dots,\, f^\mathrm{S-1}$ where $f^j$ is at
least twice the highest frequency component of any filter in the bank for the $j$th
time slice.

The time sliced templates may then be down-sampled in each interval without
aliasing, so we define them as
%
\begin{equation}
\label{eq:time-sliced-templates}
h_{i}^{s}[k] \equiv
	\begin{cases}
		h_{i}\!\left[k\frac{f}{f^s}\right] & \textrm{if } t^s \leqslant k/f^s < t^{s+1} \\
		0 & \textrm{otherwise.}
	\end{cases}
\end{equation}
%
We note that the time-slice decomposition in equation~(\ref{eq:time-slices}) is
manifestly orthogonal since the time slices are disjoint in time.  In the next
section we examine how to reduce the number of filters within each time slice
via \SVD{} of the time-sliced templates.

\begin{comment}
An example time-slice design satisfying these constraints for a $1.4 - 1.4 \,
M_{\odot}$ binary is shown in table~\ref{table:time_slices}.
%
\begin{table}[h!]
\begin{minipage}[c]{0.52\textwidth}
\centering
\vspace{0.8cm}
\includegraphics{time_slices.pdf}
\end{minipage}
\begin{minipage}[c]{0.3\textwidth}
\centering
\input{time_slices.tex}
\end{minipage}
\caption{\label{table:time_slices} Example of nearly critically sampled,
power-of-two time slices for a $1.4 - 1.4 \, M_{\odot}$ template extending from
$f_\mathrm{low} = 10 \, \mathrm{Hz}$ to $f_\mathrm{hi} = f_\mathrm{ISCO} = 1571\, \mathrm{Hz}$
with a time-frequency structure given by ($\ref{eq:fgw})$. $f^s$ is the sample
rate of the time slice, $(t^{s+1}, t^s]$ are the boundaries in seconds
preceding coalescence and \slicessamps\ is the number of samples in the
$s^{\mathrm{th}}$ filter.}
\end{table}
\end{comment}
